Normalization by Evaluation for Call-by-Push-Value and Polarized Lambda-Calculus

### 2019-02-16 fscd19/fscd19.tex
[ fscd19 ] post-submission proof reading

The idea behind \emph{normalization by evaluation} (NbE) is to utilize a
standard interpreter, usually evaluating closed terms, to compute the
normal form of an open term.
The normal form is achieved by a type-directed \emph{reification}
procedure after evaluating the open term to a semantic value, mapping
(\emph{reflecting})
the free variables to corresponding \emph{unknowns} in the semantics.
The literal use of a standard interpreter
can be achieved for the pure simply-typed lambda-calculus
\cite{bergerSchwichtenberg:lics91,filinski:semaccounttdpe}
by modelling uninterpreted base types as sets of neutral (aka atomic)
terms, or more precisely, as presheaves or sets of neutral term
families, in order to facilitate fresh bound variable generation during
reification of functions to lambdas.
Thanks to $\eta$-equality at function types, free variables of
function type can be reflected into the semantics as functions
applying the variable to their reified argument,
forming a neutral term.  This mechanism provides us with unknowns of
function type which can be faithfully reified to normal forms.
% Since neutrals are closed under application to normal arguments, they
% can act as unknown functions, storing the reified semantic argument
% they are invoked with (thus, these unknowns are aka accumulators
% \cite{gregoireleroy:icfp02}).

For the extension to sum types (logically, disjunctions and categorically,
weak coproducts), this reflection trick does not work anymore.  A
semantic value of binary sum type is either a left or a right
injection, but the decision between left or right cannot be taken at
reflection time, since a variable of sum type does not provide us with
the information.  A literal standard interpreter for closed terms can
thus no longer be used for NbE;  instead, we can utilize a monadic
interpreter.  When the interpreter attempts a case distinction on an
unknown of sum type, it asks an oracle whether the unknown is a left
or a right injection.  The oracle returns one of these alternatives,
wrapping a new unknown in the injection.  The communication with the
oracle can be modeled in a \emph{monad} $\C$,
% with the service $\tNe\,(A + B) \to \C\,(\den A + \den B)$
which records the questions asked and the continuation of the
interpreter for each of the possible answers.
A monadic semantic value is thus a \emph{case tree} where the leaves
are occupied by non-monadic values \cite{altenkirchUustalu:flops04}.
In this article we only consider weak sum types, producing non-unique
normal forms, where it does not
matter in which order the questions are asked (commuting case splits),
and whether the same
question is asked several times (redundant case splits).
The model would need refinement for strong, extensional sums
\cite{altenkirchDybjerHofmannScott:lics01,altenkirchUustalu:flops04,balatDiCosmoFiore:popl04,barral:PhD,scherer:popl17}.

@@

In contrast to the normal forms of CBN NbE, which is the algorithmic
counterpart of the completeness proof for intuitionistic propositional
logic (IPL) using Beth models, CBPV NbE gives us more restrained
normal forms, where the production of a value via injections cannot be
interrupted by more questions to the oracle.  In the research field of
focalization \cite{andreoli:focusing,liangMiller:csl07} we speak of
\emph{chaining non-invertible introductions}.
Invertible introductions are already chained in NbE thanks to
extensionality ($\eta$) for function, and more generally, negative
types.  Non-invertible eliminations are also happening in a chain,
building neutrals.  What is missing from the picture is the chaining
of invertible eliminations, i.e., case distinctions and, more
generally, pattern matching.  The picture is completed by extending
NbE to \emph{polarized lambda calculus}
\cite{zeilberger:PhD,brockNannestadSchuermann:lpar10,espiritoSanto:entcs17}
in Section~\ref{sec:fipl}.

In our presentation of the various lambda calculi we ignore the
concrete syntax, only consider the abstract syntax obtained by the
Curry-Howard-Isomorphism.  A term is simply a derivation tree whose
nodes are rule invocations.  Thus, a intrinsically typed,
nameless syntax is most natural,
and our syntactic classes are just presheaves over the category of
typing contexts and renamings.  The use of presheaves then smoothly
extents to the semantic constructions
\cite{catarina:csl93,altenkirchHofmannStreicher:ctcs95}.

@@

The category $\Cxt$ allows us to consider $A{\in}\_$ as a presheaf over
$\Cxt\op$ for any $A$, witnessed by $\treindex : \Gamma \sublist
  \Delta \to A \in \Gamma \to A \in \Delta$, which is the morphism
part of functor $A{\in}\_$ from $\Cxt$ to $\Set$, mapping object $\Gamma$ to the
set $A \in \Gamma$ of the indices of $A$ in $\Gamma$.
The associated naturality laws $\treindex\,\tid\,x \equiv x$
and $\treindex\,\tau_2\,(\treindex\,\tau_1\,x) \equiv
\treindex\,(\tau_1 \rcomp \tau_2)\,x$ hold propositionally.

@@

Simple types shall be distinguished into positive types $P$ and
negative types $N$, depending on their root type former.  Function
($\To$) and product types ($\times$ and $\tyone$) are negative,
while base types ($\tybase$) and sum types ($+$ and $\tyzero$) are positive.
\[
\begin{array}{lrl@{\qquad}l}
  A,B,C & ::= & P \mid N & \mbox{simple types} \\
  P     & ::= & \tyzero \mid A + B \mid \tybase & \mbox{positive types} \\
  N     & ::= & \tyone \mid A \times B \mid A \To B
  % P     & ::= & a \mid A + B \mid \tyzero & \mbox{positive types} \\
  % N     & ::= & A \To B \mid A \times B \mid \tyone
    & \mbox{negative types} \\
\end{array}
\]
% Simple types shall be given by the grammar
% % \fbox{$S,T ::= P \mid S \To T \mid S \times T \mid \tyone \mid S + T
% %   \mid \tyzero$}.
% \fbox{$A,B,C ::= a \mid A \To B \mid A \times B \mid \tyone \mid A + B
%   \mid \tyzero$}.
% Herein, $a$ ranges over atomic types aka base types.
Intrinsically well-typed lambda-terms, in abstract syntax,
are just inhabitants $t$ of the indexed set \fbox{$A \from \Gamma$},
inductively defined by the following rules.
\begin{gather*}
  \nru{\tvar}{A \in \Gamma}{A \from \Gamma}
\qquad
  \nru{\tabs}{B \from \cext \Gamma A}{A \To B \from \Gamma}
\qquad
  \nru{\tapp}
    {A \To B \from \Gamma \qquad A \from \Gamma}
    {B \from \Gamma}
\\[2ex]
  \nru{\tunit}{}{\tyone \from \Gamma}
\qquad
  \nru{\tpair}
    {A_1 \from \Gamma \qquad A_2 \from \Gamma}
    {A_1 \times A_2 \from \Gamma}
\qquad
  \nru{\tprj_i}
    {A_1 \times A_2 \from \Gamma}
    {A_i \from \Gamma}
\\[2ex]
  \nru{\tinj_i}
    {A_i \from \Gamma}
    {A_1 + A_2 \from \Gamma}
\qquad
  \nru{\tcase}
    {A_1 + A_2 \from \Gamma
      \qquad B \from \cext \Gamma A_1
      \qquad B \from \cext \Gamma A_2
    }{B \from \Gamma}
\qquad
  \nru{\tabort}
    {\tyzero \from \Gamma}
    {B \from \Gamma}
\end{gather*}
The skilled eye of the reader will immediately recognize the proof
rules of intuitionistic propositional logic (IPL) under the
Curry-Howard isomorphism.  Using shorthand
\fbox{$\vv n = \tvar\,(\tsuc^n\,\tzero)$} for the $n$th variable,
a term such as
\(
  \tabs\,(\tabs\,(\tpair\,
    \vv1\,
    (\tabs\,(\tapp\,\vv1\,\vv0))))
\)
could in concrete syntax be rendered as
\(
  \lambda x.\,\lambda y.\,(x,\,\lambda z.\,y\,z)
  .
\)
We leave the exact connection to a printable
syntax of the STLC to imagination of the reader,
as we shall not be concerned with considering
concrete terms in this article.

Terms of type $A$ form a presheaf
$A{\from}\_$ as witnessed by the standard weakening operation%
\footnote{Here, $\tren$ is short for \emph{renaming}, but in a nameless
calculus we should better speak of \emph{reindexing}, which could, a
bit clumsily, be also abbreviated to $\tren$.}
\fbox{$\tren : \Gamma \sublist \Delta \to A \from \Gamma \to A \from \Delta$}
defined by recursion over $t : A \from \Gamma$,
and naturality laws for $\tren$ analogously to $\treindex$.

Normal forms%
\footnote{%
There is also a stronger
notion of normal form, requiring that two \emph{extensionally equal}
lambda-terms, \ie, those that denote the
same set-theoretical function, have the same normal form
\cite{mitchell:foundations,altenkirchDybjerHofmannScott:lics01,scherer:popl17}.
Such normal forms do not have a simple inductive definition, and we
shall not consider them in this article.
}
are logically characterized as those fulfilling the
\emph{subformula} property \cite{prawitz:natded,matthes:shortproofs}.
Normal forms \fbox{$n : \tNf A\, \Gamma$} are mutually defined with
\emph{neutral} normal forms \fbox{$u : \Ne A \Gamma$}.
In the following inductive definition, we reuse the rule names from the
term constructors.
\begin{gather*}
  \nru{\tvar}{A \in \Gamma}{\Ne A \Gamma}
\qquad
  \nru{\tabs}{\NfC B {\cext \Gamma A}}{\NfP {A \To B} \Gamma}
\qquad
  \nru{\tapp}
    {\NeP {A \To B} \Gamma \qquad \Nf A \Gamma}
    {\Ne B \Gamma}
\qquad
  \nru{\tne}{\Ne \tybase \Gamma}{\Nf \tybase \Gamma}
\\[2ex]
  \nru{\tunit}{}{\Nf \tyone \Gamma}
\qquad
  \nru{\tpair}
    {\Nf {A_1} \Gamma \qquad \Nf {A_2} \Gamma}
    {\NfP {A_1 \times A_2} \Gamma}
\qquad
  \nru{\tprj_i}
    {\NeP {A_1 \times A_2} \Gamma}
    {\Ne {A_i} \Gamma}
\\[2ex]
  \nru{\tinj_i}
    {\Nf {A_i} \Gamma}
    {\NfP {A_1 + A_2} \Gamma}
\quad
  \nru{\tcase}
    {\NeP {A_1 + A_2} \Gamma
      \quad \NfC P {\cext \Gamma A_1}
      \quad \NfC P {\cext \Gamma A_2}
    }{\Nf P \Gamma}
\quad
  \nru{\tabort}
    {\Ne \tyzero \Gamma}
    {\Nf P \Gamma}
\end{gather*}
These rules only allow the elimination of neutrals; this guarantees
the subformula property and prevents any kind of computational
($\beta$) redex.
The new rule $\tne$ embeds $\tNe$ into $\tNf$,
but only at base types $\tybase$.
Further, case distinction via $\tcase$ and $\tabort$
is restricted to positive types $P$.
As a consequence, our normal forms are
\emph{$\eta$-long}, meaning that any normal inhabitant of a negative
type is a respective introduction ($\tabs$, $\tunit$, or $\tpair$).
This justifies the attribute \emph{negative} for these types:
the construction of their inhabitants proceeds mechanically, without
any choices.  In contrast, to construct an inhabitant of a
\emph{positive} type choice is required: whether case distinction is
required, and which introduction to pick in the end ($\tinj_1$ or $\tinj_2$).

Needless to say, $\tNe\,A$ and
$\tNf\,A$ are presheaves, \ie, support reindexing with $\tren$ just as terms do.
From a normal form we can extract the term via an overloaded function
\fbox{$\tm\_ : \Nf A \Gamma \to A \from \Gamma$} and
\fbox{$\tm\_ : \Ne A \Gamma \to A \from \Gamma$}
that discards constructor $\tne$ but keeps all other constructors.
This erasure function naturally commutes with reindexing, making it a
natural transformation between the presheaves $\tNf\,A$ ($\tNe\,A$,
resp.) and $A{\from}\_$.  We shall simply write, for instance,
\fbox{$\tNf\,A \todot A{\from}\_$} for such presheaf morphisms.
(The point on the arrow is mnemonic for \emph{pointwise}.)
Slightly abusive, we shall extend this notation to $n$-ary
morphisms, \eg, write $\A \todot \B \todot \C$ for
$\forall \Gamma.\ \A\,\Gamma \to (\B\,\Gamma \to \C\,\Gamma)$.

While the coproduct eliminations $\tcase$ and $\tabort$ are limited to
normal forms of positive types $P$, their extension $\tcase^B$ and
$\tabort^B$ to negative types
is admissible, for instance:
\[
\begin{array}{l@{~}lcl}
  \multicolumn 4 l {\tabort^B ~:~ \tNe\, 0 \todot \tNf\,B} \\
  \tabort^{\tyone}     & u & = & \tunit \\
  \tabort^P           & u & = & \tabort\;u \\
\end{array}
\qquad
\begin{array}{l@{~}lcl}
  \\
  \tabort^{A \times B} & u & = & \tpair\,(\tabort^A\,u)\,(\tabort^B\,u) \\
  \tabort^{A \To B}    & u & = & \tabs\,(\tabort^B\,(\tren\,\twk^A\,u)) \\
\end{array}
% \begin{array}{l@{~}lcl}
%   \multicolumn 4 l {\tabort^B ~:~ \tNe\, 0 \todot \tNf\,B} \\
%   \tabort^{\tyone}     & u & = & \tunit \\
%   \tabort^{A \times B} & u & = & \tpair\,(\tabort^A\,u)\,(\tabort^B\,u) \\
%   \tabort^{A \To B}    & u & = & \tabs\,(\tabort^B\,(\tren\,\twk^A\,u)) \\
%   \tabort^P           & u & = & \tabort\;u \\
% \end{array}
\]
$\tcase$ generalizes analogously, with a bit of care when weakening
the branches.

@@

% that does not guarantee any relation between its input term $t$ and
% output normal form $u$.
%
% \emph{Normalization by evaluation} (NbE) decomposes normalization as
% \fbox{\(
%   \norm[\Gamma] A (t : A \from \Gamma)
%   = \down[\Gamma] A \dent t {\idenv\Gamma}
% \)}
% which is
% \emph{evaluation}
% \nofbox{$\dent \_ \_ : (t : A \from \Gamma) \to \Den \Gamma \Delta \to
%   \Den A \Delta$}
% in the identity environment \nofbox{$\idenv\Gamma : \Den \Gamma \Gamma$}
% followed by
% \emph{reification}
% \nofbox{$\downG A : \Den A \Gamma \to \Nf A \Gamma$}.
\emph{Normalization by evaluation} (NbE)
\fbox{\(
%  \norm[\Gamma] A (t : A \from \Gamma)
  \tnorm\, (t : A \from \Gamma)
  = \down A \dent t {\idenv\Gamma}
\)}
decomposes normalization into
\emph{evaluation}
\nofbox{$\denty\_ : (t : A \from \Gamma) \to \den \Gamma \todot
  \den A$}
in the identity environment \nofbox{$\idenv\Gamma : \den \Gamma \Gamma$}
followed by
\emph{reification}
\nofbox{$\down A : \den A \todot \tNf A$}.
The role of evaluation is to produce from a term the corresponding
semantic (\ie, meta-theoretic) function, which is finally reified to a normal form.
Since we are evaluating open terms $t$, we need to supply an environment
$\idenv\Gamma$ which will map the free indices of $t$ to corresponding
\emph{unknowns}.  To accommodate unknowns in the semantics, types $A$
are mapped to presheaves $\den A$ (rather than just sets), and in
particular each base type $\tybase$ is mapped to the presheaf
$\tNe\,\tybase$ with the intention that the neutrals take the role of
the unknowns.  The mapping $\up A : \tNe\,A \todot \den A$ from
neutrals to unknowns is called \emph{reflection}, and defined mutually
with reification by induction on type $A$.

@@

A direct extension of our presheaf semantics to positive types cannot
work.  For instance, with $\den \tyzero = \hat0$, simply
$\up[\tyzero] \tyzero \vv0 : \bzero$ would give us an
inhabitant of the empty set, which means that reflection at the empty
type would not be definable.  Similarly, the setting
$\den{A + B} = \den A \hatplus \den B$ is refuted by
% $\up[\tyone + \tyone]^{\tyone + \tyone} \vv0 : \bone \bplus \bone$
$\up[A + B]{A + B} \vv0 : \den A (A{+}B) \bplus \den B (A{+}B)$
which would require us to make a decision of whether $A$ holds or $B$ holds
while only be given a hypothesis of type $A + B$.
Not even the usual interpretation of base types $\den o = \tNe\,o$
works in the presence of sums, as we would not be able to interpret
the term
$\tabs\,(\tcase\,\vv0\,\vv0\,\vv0) :
(o+o) \To o$ in our semantics, as $\Ne o (o+o)$ is empty.
What is needed are case distinctions on neutrals in the semantics,
allowing us the elimination of positive hypotheses before producing a
semantic value,
and we shall capture this capability in a strong monad $\C$
which can \emph{cover} the cases.

@@

For interpretation of the binders $\tabs$ and $\tcase$ we use the

@@

As already observed by Filinski
\cite[Section~5.4]{filinski:semaccounttdpe}
\cite[Section~3.2]{filinski:tlca01},
normalization by evaluation can be carried out in the continuation
monad.
In our setting, we use a continuation monad $\CC$ on presheaves
defined as
% An alternative implementation of $\C$ is the continuation monad $\CC$
% on presheaves as, for instance, employed by Ilik et al.\
% \cite{ilikLeeHerbelin:apal10}
% for Kripke models and NbE of classical logic.
\[
  \CC\;\J = \forall A.\ (\J \hatto \tNf\,A) \hatto \tNf\,A
  .
\]
The answer type of this continuation monad is always $\tNF$, however,
we are polymorphic in the simple type $A$ of normal forms we produce.%

>>>>

The idea behind \emph{normalization by evaluation} (NbE) is to utilize a
standard interpreter, usually evaluating closed terms, to compute the
normal form of an open term.
The normal form is obtained by a type-directed \emph{reification}
procedure after evaluating the open term to a semantic value, mapping
(\emph{reflecting})
the free variables to corresponding \emph{unknowns} in the semantics.
The literal use of a standard interpreter
can be achieved for the pure simply-typed lambda-calculus
\cite{bergerSchwichtenberg:lics91,filinski:semaccounttdpe}
by modelling uninterpreted base types as sets of neutral (aka atomic)
terms, or more precisely, as presheaves or sets of neutral term
families, in order to facilitate fresh bound variable generation during
reification of functions to lambdas.
Thanks to $\eta$-equality at function types, free variables of
function type can be reflected into the semantics as functions
applying the variable to their reified argument,
forming a neutral term.  This mechanism provides us with unknowns of
function type which can be faithfully reified to normal forms.
% Since neutrals are closed under application to normal arguments, they
% can act as unknown functions, storing the reified semantic argument
% they are invoked with (thus, these unknowns are aka accumulators
% \cite{gregoireleroy:icfp02}).

For the extension to sum types (logically, disjunctions, and categorically,
weak coproducts), this reflection trick does not work anymore.  A
semantic value of binary sum type is either a left or a right
injection, but the decision between left or right cannot be taken at
reflection time, since a variable of sum type does not provide us with
such information.  A literal standard interpreter for closed terms can
thus no longer be used for NbE;  instead, we can utilize a monadic
interpreter.  When the interpreter attempts a case distinction on an
unknown of sum type, it asks an oracle whether the unknown is a left
or a right injection.  The oracle returns one of these alternatives,
wrapping a new unknown in the respective injection.  The communication with the
oracle can be modeled in a \emph{monad} $\C$,
% with the service $\tNe\,(A + B) \to \C\,(\den A + \den B)$
which records the questions asked and the continuation of the
interpreter for each of the possible answers.
A monadic semantic value is thus a \emph{case tree} where the leaves
are occupied by non-monadic values \cite{altenkirchUustalu:flops04}.
In this article we only consider weak sum types, producing non-unique
normal forms, where it does not
matter in which order the questions are asked (commuting case splits),
and whether the same
question is asked several times (redundant case splits).
The model would need refinement for strong, extensional sums
\cite{altenkirchDybjerHofmannScott:lics01,altenkirchUustalu:flops04,balatDiCosmoFiore:popl04,barral:PhD,scherer:popl17}.

@@

In contrast to the normal forms of CBN NbE, which is the algorithmic
counterpart of the completeness proof for intuitionistic propositional
logic (IPL) using Beth models, CBPV NbE gives us more restrained
normal forms, where the production of a value via injections cannot be
interrupted by more questions to the oracle.  In the research field of
focalization \cite{andreoli:focusing,liangMiller:csl07} we speak of
\emph{chaining non-invertible introductions}.
Invertible introductions are already chained in NbE thanks to
extensionality ($\eta$) for function, and more generally, negative
types.  Non-invertible eliminations are also happening in a chain when
building neutrals.  What is missing from the picture is the chaining
of invertible eliminations, i.e., case distinctions and, more
generally, pattern matching.  The picture is completed by extending
NbE to \emph{polarized lambda calculus}
\cite{zeilberger:PhD,brockNannestadSchuermann:lpar10,espiritoSanto:entcs17}
in Section~\ref{sec:fipl}.

In our presentation of the various lambda calculi we ignore the
concrete syntax, only consider the abstract syntax obtained by the
Curry-Howard-Isomorphism.  A term is simply a derivation tree whose
nodes are rule invocations.  Thus, a intrinsically typed,
nameless syntax is most natural,
and our syntactic classes are all presheaves over the category of
typing contexts and renamings.  The use of presheaves then smoothly
extents to the semantic constructions
\cite{catarina:csl93,altenkirchHofmannStreicher:ctcs95}.

@@

The category $\Cxt$ allows us to consider $A{\in}\_$ as a presheaf over
$\Cxt\op$ for any $A$, witnessed by $\treindex : \Gamma \sublist
  \Delta \to A \in \Gamma \to A \in \Delta$, which is the morphism
part of functor $A{\in}\_$ from $\Cxt$ to $\Set$, mapping object $\Gamma$ to the
set $A \in \Gamma$ of the indices of $A$ in $\Gamma$.
The associated functor laws $\treindex\,\tid\,x \equiv x$
and $\treindex\,\tau_2\,(\treindex\,\tau_1\,x) \equiv
\treindex\,(\tau_1 \rcomp \tau_2)\,x$ hold propositionally.

@@

Simple types shall be distinguished into positive types $P$ and
negative types $N$, depending on their root type former.  Function
($\To$) and product types ($\times$ and $\tyone$) are negative,
while base types ($\tybase$) and sum types ($+$ and $\tyzero$) are positive.
\[
\begin{array}{lrl@{\qquad}l}
  A,B,C & ::= & P \mid N & \mbox{simple types} \\
  P     & ::= & \tyzero \mid A + B \mid \tybase & \mbox{positive types} \\
  N     & ::= & \tyone \mid A \times B \mid A \To B
  % P     & ::= & a \mid A + B \mid \tyzero & \mbox{positive types} \\
  % N     & ::= & A \To B \mid A \times B \mid \tyone
    & \mbox{negative types} \\
\end{array}
\]
% Simple types shall be given by the grammar
% % \fbox{$S,T ::= P \mid S \To T \mid S \times T \mid \tyone \mid S + T
% %   \mid \tyzero$}.
% \fbox{$A,B,C ::= a \mid A \To B \mid A \times B \mid \tyone \mid A + B
%   \mid \tyzero$}.
% Herein, $a$ ranges over atomic types aka base types.
Intrinsically well-typed lambda-terms, in abstract syntax,
are just inhabitants $t$ of the indexed set \fbox{$A \from \Gamma$},
inductively defined by the following rules.
\begin{gather*}
  \nru{\tvar}{A \in \Gamma}{A \from \Gamma}
\qquad
  \nru{\tabs}{B \from \cext \Gamma A}{A \To B \from \Gamma}
\qquad
  \nru{\tapp}
    {A \To B \from \Gamma \qquad A \from \Gamma}
    {B \from \Gamma}
\\[2ex]
  \nru{\tunit}{}{\tyone \from \Gamma}
\qquad
  \nru{\tpair}
    {A_1 \from \Gamma \qquad A_2 \from \Gamma}
    {A_1 \times A_2 \from \Gamma}
\qquad
  \nru{\tprj_i}
    {A_1 \times A_2 \from \Gamma}
    {A_i \from \Gamma}
\\[2ex]
  \nru{\tinj_i}
    {A_i \from \Gamma}
    {A_1 + A_2 \from \Gamma}
\qquad
  \nru{\tcase}
    {A_1 + A_2 \from \Gamma
      \qquad B \from \cext \Gamma A_1
      \qquad B \from \cext \Gamma A_2
    }{B \from \Gamma}
\qquad
  \nru{\tabort}
    {\tyzero \from \Gamma}
    {B \from \Gamma}
\end{gather*}
The skilled eye of the reader will immediately recognize the proof
rules of intuitionistic propositional logic (IPL) under the
Curry-Howard isomorphism, where $A \from \Gamma$ is to be read as
``$A$ follows from $\Gamma$''.  Using shorthand
\fbox{$\vv n = \tvar\,(\tsuc^n\,\tzero)$} for the $n$th variable,
a term such as
\(
  \tabs\,(\tabs\,(\tpair\,
    \vv1\,
    (\tabs\,(\tapp\,\vv1\,\vv0))))
\)
could in concrete syntax be rendered as
\(
  \lambda x.\,\lambda y.\,(x,\,\lambda z.\,y\,z)
  .
\)
We leave the exact connection to a printable
syntax of the STLC to the imagination of the reader,
as we shall not be concerned with considering
concrete terms in this article.

Terms of type $A$ form a presheaf
$A{\from}\_$ as witnessed by the standard weakening operation%
\footnote{Here, $\tren$ is short for \emph{renaming}, but in a nameless
calculus we should better speak of \emph{reindexing}, which could, a
bit clumsily, be also abbreviated to $\tren$.}
\fbox{$\tren : \Gamma \sublist \Delta \to A \from \Gamma \to A \from \Delta$}
defined by recursion over $t : A \from \Gamma$,
and functor laws for $\tren$ analogously to $\treindex$.

Normal forms%
\footnote{%
There is also a stronger
notion of normal form, requiring that two \emph{extensionally equal}
lambda-terms, \ie, those that denote the
same set-theoretical function, have the same normal form
\cite{mitchell:foundations,altenkirchDybjerHofmannScott:lics01,scherer:popl17}.
Such normal forms do not have a simple inductive definition, and we
shall not consider them in this article.
}
are logically characterized as those fulfilling the
\emph{subformula} property \cite{prawitz:natded,matthes:shortproofs}.
Normal forms \fbox{$n : \tNf A\, \Gamma$} are mutually defined with
\emph{neutral} normal forms \fbox{$u : \Ne A \Gamma$}.
In the following inductive definition, we reuse the rule names from the
term constructors.
\begin{gather*}
  \nru{\tvar}{A \in \Gamma}{\Ne A \Gamma}
\qquad
  \nru{\tabs}{\NfC B {\cext \Gamma A}}{\NfP {A \To B} \Gamma}
\qquad
  \nru{\tapp}
    {\NeP {A \To B} \Gamma \qquad \Nf A \Gamma}
    {\Ne B \Gamma}
\qquad
  \nru{\tne}{\Ne \tybase \Gamma}{\Nf \tybase \Gamma}
\\[2ex]
  \nru{\tunit}{}{\Nf \tyone \Gamma}
\qquad
  \nru{\tpair}
    {\Nf {A_1} \Gamma \qquad \Nf {A_2} \Gamma}
    {\NfP {A_1 \times A_2} \Gamma}
\qquad
  \nru{\tprj_i}
    {\NeP {A_1 \times A_2} \Gamma}
    {\Ne {A_i} \Gamma}
\\[2ex]
  \nru{\tinj_i}
    {\Nf {A_i} \Gamma}
    {\NfP {A_1 + A_2} \Gamma}
\quad
  \nru{\tcase}
    {\NeP {A_1 + A_2} \Gamma
      \quad \NfC P {\cext \Gamma A_1}
      \quad \NfC P {\cext \Gamma A_2}
    }{\Nf P \Gamma}
\quad
  \nru{\tabort}
    {\Ne \tyzero \Gamma}
    {\Nf P \Gamma}
\end{gather*}
These rules only allow the elimination of \emph{neutrals}; this
restriction guarantees
the subformula property and prevents any kind of computational
($\beta$) redex.
The new rule $\tne$ embeds $\tNe$ into $\tNf$,
but only at base types $\tybase$ \cite[Section~3.3]{altenkirchHofmannStreicher:ctcs95}.
Further, case distinction via $\tcase$ and $\tabort$
is restricted to positive types $P$.
As a consequence, our normal forms are
\emph{$\eta$-long}, meaning that any normal inhabitant of a negative
type is a respective introduction ($\tabs$, $\tunit$, or $\tpair$).
This justifies the attribute \emph{negative} for these types:
the construction of their inhabitants proceeds mechanically, without
any choices.  In contrast, constructing an inhabitant of a
\emph{positive} type involves choice: whether case distinction is
required, and which introduction to pick in the end ($\tinj_1$ or $\tinj_2$).

Needless to say, $\tNe\,A$ and
$\tNf\,A$ are presheaves, \ie, support reindexing with $\tren$ just as terms do.
From a normal form we can extract the term via an overloaded function
\fbox{$\tm\_ : \Nf A \Gamma \to A \from \Gamma$} and
\fbox{$\tm\_ : \Ne A \Gamma \to A \from \Gamma$}
that discards constructor $\tne$ but keeps all other constructors.
This erasure function naturally commutes with reindexing, making it a
natural transformation between the presheaves $\tNf\,A$ ($\tNe\,A$,
resp.) and $A{\from}\_$.  We shall simply write, for instance,
$\tNf\,A \todot A{\from}\_$ for such presheaf morphisms.
(The point on the arrow is mnemonic for \emph{pointwise}.)
Slightly abusive, we shall extend this notation to $n$-ary
morphisms, \eg, write $\A \todot \B \todot \C$ for
$\forall \Gamma.\ \A\,\Gamma \to (\B\,\Gamma \to \C\,\Gamma)$.

\begin{remark*}
While the coproduct eliminations $\tcase$ and $\tabort$ are limited to
normal forms of positive types $P$, their extension $\tcase^B$ and
$\tabort^B$ to negative types
is admissible, for instance:
\[
\begin{array}{l@{~}lcl}
  \multicolumn 4 l {\tabort^B ~:~ \tNe\, 0 \todot \tNf\,B} \\
  \tabort^{\tyone}     & u & = & \tunit \\
  \tabort^P           & u & = & \tabort\;u \\
\end{array}
\qquad
\begin{array}{l@{~}lcl}
  \\
  \tabort^{A \times B} & u & = & \tpair\,(\tabort^A\,u)\,(\tabort^B\,u) \\
  \tabort^{A \To B}    & u & = & \tabs\,(\tabort^B\,(\tren\,\twk^A\,u)) \\
\end{array}
% \begin{array}{l@{~}lcl}
%   \multicolumn 4 l {\tabort^B ~:~ \tNe\, 0 \todot \tNf\,B} \\
%   \tabort^{\tyone}     & u & = & \tunit \\
%   \tabort^{A \times B} & u & = & \tpair\,(\tabort^A\,u)\,(\tabort^B\,u) \\
%   \tabort^{A \To B}    & u & = & \tabs\,(\tabort^B\,(\tren\,\twk^A\,u)) \\
%   \tabort^P           & u & = & \tabort\;u \\
% \end{array}
\]
$\tcase$ generalizes analogously, with a bit of care when weakening
the branches.
\end{remark*}

@@

% that does not guarantee any relation between its input term $t$ and
% output normal form $u$.
%
% \emph{Normalization by evaluation} (NbE) decomposes normalization as
% \fbox{\(
%   \norm[\Gamma] A (t : A \from \Gamma)
%   = \down[\Gamma] A \dent t {\idenv\Gamma}
% \)}
% which is
% \emph{evaluation}
% \nofbox{$\dent \_ \_ : (t : A \from \Gamma) \to \Den \Gamma \Delta \to
%   \Den A \Delta$}
% in the identity environment \nofbox{$\idenv\Gamma : \Den \Gamma \Gamma$}
% followed by
% \emph{reification}
% \nofbox{$\downG A : \Den A \Gamma \to \Nf A \Gamma$}.
\emph{Normalization by evaluation} (NbE)
\fbox{\(
%  \norm[\Gamma] A (t : A \from \Gamma)
  \tnorm\, (t : A \from \Gamma)
  = \down A \dent t {\idenv\Gamma}
\)}
decomposes normalization into
\emph{evaluation}
\nofbox{$\denty\_ : (t : A \from \Gamma) \to \den \Gamma \todot
  \den A$}
in the identity environment \nofbox{$\idenv\Gamma : \den \Gamma \Gamma$}
followed by
\emph{reification}
\nofbox{$\down A : \den A \todot \tNf A$}
(aka quoting).
The role of evaluation is to produce from a term the corresponding
semantic (\ie, meta-theoretic) function, which is finally reified to a normal form.
Since we are evaluating open terms $t$, we need to supply an environment
$\idenv\Gamma$ which will map the free indices of $t$ to corresponding
\emph{unknowns}.  To accommodate unknowns in the semantics, types $A$
are mapped to presheaves $\den A$ (rather than just sets), and in
particular each base type $\tybase$ is mapped to the presheaf
$\tNe\,\tybase$ with the intention that the neutrals take the role of
the unknowns.  The mapping $\up A : \tNe\,A \todot \den A$ from
neutrals to unknowns is called \emph{reflection} (aka unquoting), and defined mutually
with reification by induction on type $A$.

@@

A direct extension of our presheaf semantics to positive types cannot
work.  For instance, with $\den \tyzero = \hat0$, simply
% $\up[\tyzero] \tyzero \vv0 : \bzero$
$\tfresh^{\tyzero}_{\cempty} : \bzero$
would give us an
inhabitant of the empty set, which means that reflection at the empty
type would not be definable.  Similarly, the setting
$\den{A + B} = \den A \hatplus \den B$ is refuted by
% $\up[\tyone + \tyone]^{\tyone + \tyone} \vv0 : \bone \bplus \bone$
$\tfresh^{A + B}_{\cempty} : \den A (A{+}B) \bplus \den B (A{+}B)$
% $\up[A + B]{A + B} \vv0 : \den A (A{+}B) \bplus \den B (A{+}B)$
which would require us to make a decision of whether $A$ holds or $B$ holds
while only be given a hypothesis of type $A + B$.
Not even the usual interpretation of base types $\den o = \tNe\,o$
works in the presence of sums, as we would not be able to interpret
the term
$\tabs\,(\tcase\,\vv0\,\vv0\,\vv0) :
(o+o) \To o$ in our semantics, as $\Ne o (o+o)$ is empty.
What is needed are case distinctions on neutrals in the semantics,
allowing us the elimination of positive hypotheses before producing a
semantic value,
and we shall capture this capability in a strong monad $\C$
which can \emph{cover} the cases.

@@

For the interpretation of the binders $\tabs$ and $\tcase$ we use the

@@

As already observed by Filinski
\cite[Section~5.4]{filinski:semaccounttdpe}
\cite[Section~3.2]{filinski:tlca01},
normalization by evaluation can be carried out in the continuation
monad.
In our setting, we use a continuation monad $\CC$ on presheaves
defined as
% An alternative implementation of $\C$ is the continuation monad $\CC$
% on presheaves as, for instance, employed by Ilik et al.\
% \cite{ilikLeeHerbelin:apal10}
% for Kripke models and NbE of classical logic.
\[
  \CC\;\J = \forall A.\ (\J \hatto \tNf\,A) \hatto \tNf\,A
  .
\]
The answer type of this continuation monad is always $\tNf$, however,
we are polymorphic in the simple type $A$ of normal forms we produce.%
